---
title: Final lab survival analysis
author: "AS Huffmyer"
date: '2026'
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_depth: 6
    toc_float: yes
  pdf_document:
    keep_tex: yes
editor_options: 
  chunk_output_type: console
---

# Set up 

Set up workspace, set options, and load required packages.    
```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries. 
```{r}
library(tidyverse)
library(ggplot2)
library(survival)
library(readxl)
library(ggsurvfit)
library(gtsummary)
library(cardx)
library(cowplot)
library(lme4)
library(lmerTest)
library(car)
library(emmeans)
```

# Read in data 

Read in data. 
```{r}
data<-read_csv("data/final-lab-survival/final_lab_survival.csv")%>%
  mutate(time=as.numeric(time))
```

Read in metadata. 
```{r}
metadata<-read_csv("data/final-lab-survival/final-lab-survival-metadata.csv")%>%select(!c(notes))
```

Merge in metadata. 
```{r}
data<-left_join(data, metadata)
head(data)
```

"group" here refers to the original hardening treatment. All oysters were run in the survival assay at 34-36°C.  

# Plot temperatures 

Plot temperatures for each round 

```{r}
temps<-read_csv("data/final-lab-survival/temperature.csv")
```

Set theme. 
```{r}
my_theme<-theme_classic()+
  theme(
    axis.title.y = element_text(size = 14, color = "black"),
    axis.text.y = element_text(size=12, color = "black"),
    axis.text.x = element_text(size=12, color = "black"),
    axis.title.x =element_text(size = 14, color = "black"),
    legend.position = "none")
```

Plot faceted by round. 

```{r}
plot<-temps%>%
  
  ggplot(aes(x=time, y=temp.C, color=round))+
  facet_grid(~round)+
  geom_point()+
  stat_summary(fun = mean, geom = "line", linewidth = 1)+
  my_theme+
  theme(legend.position="right");plot

ggsave(filename="figures/final-lab-survival/temperature_rounds.png", width=10, height=5)
```

Reset levels.
```{r}
data$group<-relevel(as.factor(data$group), ref="10K Control")
```

# Generate survival curves

```{r}
s1 <- survival::survfit(Surv(time, status) ~ group, data = data)
str(s1)
```

Plot the survival function
```{r}
ggsurvfit::survfit2(Surv(time, status) ~ group, data = data) %>% 
  ggsurvfit::ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability"
  )
```

Estimate survival at 48 hours. 
```{r}
summary(survival::survfit(Surv(time, status) ~ group, data = data), times = 48)
```

Use a log rank model to determine statistical differences in curves. 
```{r}
survival::survdiff(Surv(time, status) ~ group, data = data)
```

Analyze again with a Cox proportional hazards model. 
```{r}
survival::coxph(Surv(time, status) ~ group, data = data)

survival::coxph(Surv(time, status) ~ group, data = data) %>% 
  tbl_regression(exp = TRUE) 
```

## Generate plots 

### Treatment 

Round 1
```{r}
plot1<-ggsurvfit::survfit2(Surv(time, status) ~ group, data = data%>%filter(round=="round1")) %>% 
  ggsurvfit::ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title=""
  )+
  ylim(0,1)+
  my_theme+
  #scale_linetype_manual(values=c("solid", "dashed"))+
  #scale_colour_manual(values=c("darkgray", "orange"))+
  theme(legend.position="right");plot1
```

Round 2

All rounds

# Run binomial logistic models 

Run a binomial logistic model to account for variation between rounds. Include round and oyster size (length in mm) as random effects. 

0 = alive
1 = mortality 

```{r}
#surv_model<-glmer(status ~ time * group + (1|round) + (1|length.mm), family = binomial(link = "logit"), data=data)
surv_model<-glmer(status ~ time * group + (1|length.mm), family = binomial(link = "logit"), data=data)

summary(surv_model)
Anova(surv_model)

emm<-emmeans(surv_model, ~group)
pairs(emm)

```

Evaluate random effects of round and length (mm). 
```{r}
# Plot the random effects
ranef(surv_model)

library(sjPlot)

plot_model(surv_model, type = "re")

library(performance)
r2(surv_model)
```

Generate predictive mortality plot. 

```{r}
# Generate predicted probabilities
data$predicted_mortality <- predict(surv_model, type = "response")

# Plot
plot2<-ggplot(data, aes(x = time, y = predicted_mortality, color = group, fill = group, group=group)) +
  geom_point(aes(y = status), alpha = 0.6, position = position_jitter(height = 0.03)) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), se = FALSE) +
  #scale_color_manual(values=c("cyan3", "coral"))+
  labs(
    title = "",
    y = "Probability of Mortality",
  ) +
  theme_classic();plot2

ggsave(plot2, filename="figures/final-lab-survival/survival_group.png", width=8, height=6)
```

# LT50 Analysis

Calculate the time of 50% mortality (LT50) for each treatment from the binomial model. LT50 is solved by finding where the model's predicted probability of mortality = 0.5, using population-level predictions (random effects for round excluded via `re.form = NA`).

```{r lt50-table}
# Function to find LT50 for a given family and treatment using uniroot
# re.form = NA excludes round-level random effects (population-level estimate)
get_lt50 <- function(model, group_val, time_range = c(0, 300)) {
  pred_fn <- function(t) {
    newdata <- data.frame(
      time      = t,
      group = group_val
    )
    predict(model, newdata = newdata, type = "response", re.form = NA) - 0.5
  }

  # Verify the predicted curve actually crosses 0.5 within the search range
  lo <- tryCatch(pred_fn(time_range[1]), error = function(e) NA_real_)
  hi <- tryCatch(pred_fn(time_range[2]), error = function(e) NA_real_)
  if (anyNA(c(lo, hi)) || sign(lo) == sign(hi)) return(NA_real_)

  tryCatch(
    uniroot(pred_fn, interval = time_range)$root,
    error = function(e) NA_real_
  )
}
```

```{r}
# Compute LT50 for every family × treatment combination
lt50_results <- data %>%
  distinct(group) %>%
  rowwise() %>%
  mutate(
    LT50_hrs = get_lt50(surv_model, as.character(group))
  ) %>%
  ungroup()

# Identify the non-control treatment label automatically
treated_label <- setdiff(unique(data$group), "10K Control")

# Pivot wide so control and treated LT50 sit side by side per family
lt50_table <- lt50_results %>%
  pivot_wider(names_from = group, values_from = LT50_hrs) %>%
  rename(
    LT50_control  = `10K Control`,
    LT50_35C  = `10K 35C`,
    LT50_FW  = `10K FW`,
    LT50_PolyIC  = `10K PolyIC`,
    LT50_FW35C  = `10K FW 35C`,
  ) %>%
  mutate(
    LT50_control  = round(LT50_control,  1),
    LT50_35C  = round(LT50_35C,  1),
    LT50_FW  = round(LT50_FW,  1),
    LT50_PolyIC  = round(LT50_PolyIC,  1),
    LT50_FW35C  = round(LT50_FW35C,  1)
  ) 

knitr::kable(
  lt50_table
)
```  