---
title: "Viewing logger files"
author: "Ariana S Huffmyer"
date: "2024"
output: 
  html_document:
    code_folding: hide
    toc: yes
    toc_depth: 6
    toc_float: yes
editor_options: 
  chunk_output_type: console
---

This script reads and plots environmental data from Hobo loggers. 

# Set Up    

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

```{r}
library(tidyverse)
library(stringr)
library(readxl)
library(purrr)
library(lubridate)
library(ggplot2)
library(seacarb)
library(broom)
library(cowplot)
```

# Point Whitney 

## Read in files 

Read in Hobo Tidbit files  
```{r}
tidbit.files <- list.files(path="data/loggers", pattern="*.xlsx", full.names=TRUE)

#will need to change to HST time 
tidbits <- setNames(tidbit.files, tidbit.files) %>%
    map_dfr(read_xlsx, .id = "logger") %>%
    rename(DateTime=`Date-Time (PDT)`, temp.C=`Temperature , 째C`)%>%
    select(!`#`)%>%
    select(logger, DateTime, temp.C)%>%
    mutate(logger=sub(".*/", "", logger))%>% #remove the path name
    mutate(logger=sub("_.*", "", logger))%>% #keep only the serial number 
    filter(!temp.C=="NA")

tidbits$DateTime<-as.POSIXct(tidbits$DateTime, format="%m/%d/%Y %H:%M:%S")

```

List the logger serial numbers in this dataframe.  
```{r}
levels(as.factor(tidbits$logger))
```

There are 2 loggers as expected.  

Add in group information. 

```{r}
tidbits <- tidbits %>%
  mutate(group = c("10K Dock"))

levels(as.factor(tidbits$group))
```

Group information is now added.  

## Plot data 

All groups 
```{r}
temp_plot<-tidbits%>%
  
  ggplot(aes(x=DateTime, y=temp.C))+
  geom_point(alpha=0.7, color="darkgray")+
  geom_smooth(color="black")+
  ylab("Temperature (째C)")+
  xlab("Date")+
  ggtitle("10K Lagoon Dock")+
  ylim(0,32)+
  theme_classic(); temp_plot

ggsave(temp_plot, filename="figures/environmental/10K-loggers.png", width=6, height=4)
```

In the lagoon, temperatures reached a max of about 25-26C in late July with a minimum around 15C.  
# Baywater Outplant 

## Read in files 

Read in Hobo Tidbit files  
```{r}
tidbit.files <- list.files(path="data/loggers/outplant", pattern="*.xlsx", full.names=TRUE)

#will need to change to HST time 
tidbits <- setNames(tidbit.files, tidbit.files) %>%
    map_dfr(read_xlsx, .id = "logger") %>%
    rename(DateTime=`Date-Time (PDT)`, temp.C=`Temp , 째C`)%>%
    select(!`#`)%>%
    select(logger, DateTime, temp.C)%>%
    mutate(logger=sub(".*/", "", logger))%>% #remove the path name
    mutate(logger=sub("_.*", "", logger))%>% #keep only the serial number 
    filter(!temp.C=="NA")

tidbits$DateTime<-as.POSIXct(tidbits$DateTime, format="%m/%d/%Y %H:%M:%S")

tidbits%>%write_csv(file="output/baywater_temperature_20250822.csv")

```

List the logger serial numbers in this dataframe.  
```{r}
levels(as.factor(tidbits$logger))
```

There are 3 loggers as expected.  

Add in group information. 

```{r}
tidbits <- tidbits %>%
  mutate(group = c("10K Outplant"))

levels(as.factor(tidbits$group))
```

Group information is now added.  

## Plot data 

All groups 
```{r}
temp_plot<-tidbits%>%
  
  ggplot(aes(x=DateTime, y=temp.C))+
  geom_point(alpha=0.7, color="darkgray", size=1)+
  geom_smooth(color="black")+
  ylab("Temperature (째C)")+
  xlab("Date")+
  ggtitle("10K Outplants - Baywater")+
  ylim(-5,32)+
  theme_classic(); temp_plot

ggsave(temp_plot, filename="figures/environmental/10K-outplant-loggers.png", width=6, height=4)

```

