---
title: "10K seed outplant growth analysis"
author: "AS Huffmyer"
date: '2025'
output:
  md_document: default
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 6
    toc_float: true
  github_document: null
editor_options:
  chunk_output_type: console
---

I still need to load in initial sizes, this analysis currently examines final size, not normalized to initial size. 

# Set up 

Set up workspace, set options, and load required packages.    
```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries. 
```{r}
if ("tidyverse" %in% rownames(installed.packages()) == 'FALSE') install.packages('tidyverse') 
if ("ggplot2" %in% rownames(installed.packages()) == 'FALSE') install.packages('ggplot2') 
if ("ggeffects" %in% rownames(installed.packages()) == 'FALSE') install.packages('ggeffects') 

library("ggplot2")
library("tidyverse")
library("lme4")
library("lmerTest")
library("emmeans")
library("car")
library("mgcv")
library("ggeffects")
library("cowplot")
library("readxl")
```

# Load data 

Read in data files. 
```{r}
data<-read_csv("data/outplant/outplant-growth.csv")%>%
  mutate(date=as.factor(date))
```

Set data attributes. 
```{r}
data$purple.tag<-factor(data$purple.tag)
```

Add in treatment information 
```{r}
metadata<-read_csv("data/outplant/outplant-metadata.csv")
metadata$purple.tag<-factor(metadata$purple.tag)

data$treatment<-metadata$treatment[match(data$purple.tag, metadata$purple.tag)]
data$source.bag<-metadata$source.bag[match(data$purple.tag, metadata$purple.tag)]
```

# Plot data 

Plot a histogram of each data column to identify any outliers. 
```{r}
hist(data$length.mm)
hist(data$width.mm)
```

No obvious outliers. 

Plot a box plot of length and width as a function of bag number, colored by treatment. 

```{r}
plot1<-data%>%
  ggplot(aes(x=purple.tag, y=length.mm, fill = treatment)) +
  facet_wrap(~date)+
  geom_boxplot(width=.5, outlier.shape= NA, position = position_dodge(width = 0.4)) +
  geom_point(pch = 21, position=position_jitterdodge(dodge.width=0.4)) +
  xlab("Bag") + 
  theme_classic(); plot1
```

Bag 12 is an outlier in the field and was much larger than all others. Remove for analysis. 

```{r}
data<-data%>%
  filter(!purple.tag=="12")
```

Plot again
```{r}
plot1<-data%>%
  ggplot(aes(x=purple.tag, y=length.mm, fill = treatment)) +
  facet_wrap(~date)+
  geom_boxplot(width=.5, outlier.shape= NA, position = position_dodge(width = 0.4)) +
  geom_point(pch = 21, position=position_jitterdodge(dodge.width=0.4)) +
  xlab("Bag") + 
  theme_classic(); plot1
```

Plot width. 
```{r}
plot2<-data%>%
  ggplot(aes(x=purple.tag, y=width.mm, fill = treatment)) +
  facet_wrap(~date)+
  geom_boxplot(width=.5, outlier.shape= NA, position = position_dodge(width = 0.4)) +
  geom_point(pch = 21, position=position_jitterdodge(dodge.width=0.4)) +
  xlab("Bag") + 
  theme_classic(); plot2
```

# Generate a metric of volume

Use the equation to predict depth from length and width using growth data from Goose Point from the same source population. Use to calculate volume from length and width measured here.  

Obtain training data with known length width and depth from 20240909 dataset. 
```{r}
training_data<-read_csv(file="https://raw.githubusercontent.com/RobertsLab/project-gigas-conditioning/refs/heads/main/data/outplanting/GoosePoint/growth_GoosePoint.csv")%>%
  mutate(date=as.factor(date))%>%
  filter(date %in% c("20240909", "20251009"))

training_data$volume <- (4/3) * pi * (training_data$length.mm/2) * (training_data$width.mm/2) * (training_data$depth.mm/2)

training_data<-training_data%>%
  filter(!is.na(volume))

hist(training_data$volume)
```

Fit a polynomial regression and a GAM model and compare the fits. 
```{r}
# Fit a polynomial regression model (2nd-degree polynomial)
poly_model <- lm(volume ~ poly(length.mm, 2) + poly(width.mm, 2) + length.mm:width.mm, data = training_data)

summary(poly_model)
```

Plot predicted vs actual volume from training data. 
```{r}
training_data$Predicted_Volume_Poly <- predict(poly_model, training_data)
plot(training_data$Predicted_Volume_Poly ~ training_data$volume)
```

Predict volume for our test data set. 

```{r}
data$Predicted_Volume_Poly <- predict(poly_model, data)
hist(data$Predicted_Volume_Poly)
```

Remove any observation that resulted in a negative value. 

```{r}
data<-data%>%filter(Predicted_Volume_Poly>0)
hist(data$Predicted_Volume_Poly)
```

Proceed with the calculated volume data.  

Remove upper outliers. 

```{r}
data<-data%>%
  filter(Predicted_Volume_Poly<125000)
```

# Analyze volume for each treatment

## Plot data 

```{r}
plot4<-data%>%
  ggplot(aes(x=purple.tag, y=Predicted_Volume_Poly, fill = treatment)) +
  geom_boxplot(width=.5, outlier.shape= NA, position = position_dodge(width = 0.4)) +
  geom_point(pch = 21, position=position_jitterdodge(dodge.width=0.4)) +
  xlab("Bag") + 
  theme_classic(); plot4

ggsave(plot4, filename="figures/growth/bag_sizes.png", width=8, height=5, units="in")
```

Plot summarized by treatment 

```{r}
plot5<-data%>%
  ggplot(aes(x=treatment, y=Predicted_Volume_Poly, fill = treatment)) +
  geom_boxplot(width=.5, outlier.shape= NA, position = position_dodge(width = 0.4)) +
  geom_point(pch = 21, position=position_jitterdodge(dodge.width=0.4)) +
  xlab("Treatment") + 
  theme_classic() +
  theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1)); plot5

ggsave(plot5, filename="figures/growth/treatment_sizes.png", width=6, height=4, units="in")
```

No clear differences in treatment. 

## Run linear mixed models models 

```{r}
model<-lmer(sqrt(Predicted_Volume_Poly) ~ treatment + (1|purple.tag:treatment), data=data)

summary(model)
anova(model)

qqPlot(residuals(model))
hist(residuals(model))
```

No difference by treatment. 

There might be several outliers. Conduct outlier removal and analyze again.  

```{r}
# Extract raw residuals
data$raw_resid <- residuals(model)

# Standardize residuals
data$std_resid <- data$raw_resid / sd(data$raw_resid)

# Flag potential outliers
outlier_threshold <- 3
data$outlier_flag <- abs(data$std_resid) > outlier_threshold

# Filter rows flagged as outliers
outliers <- data %>% filter(outlier_flag == TRUE)
print(outliers)

# Plot standardized residuals
ggplot(data, aes(x = seq_along(std_resid), y = std_resid)) +
  geom_point(aes(color = outlier_flag), size = 2) +
  geom_hline(yintercept = c(-outlier_threshold, outlier_threshold), linetype = "dashed", color = "red") +
  labs(title = "Standardized Residuals with Outliers", x = "Index", y = "Standardized Residual") +
  theme_minimal()
```
No outliers. 

Remove identified outliers. 
```{r}
data<-data%>%
  filter(!outlier_flag==TRUE)
```

Analyze again. 

```{r}
model<-lmer(sqrt(Predicted_Volume_Poly) ~ treatment + (1|purple.tag:treatment), data=data)

summary(model)
anova(model)

qqPlot(residuals(model))
hist(residuals(model))
```

No treatment effects.  