---
title: 10K seed resazurin metabolic rate analysis
author: "AS Huffmyer"
date: '2024'
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_depth: 6
    toc_float: yes
  pdf_document:
    keep_tex: yes
editor_options: 
  chunk_output_type: console
---

# Set up 

Set up workspace, set options, and load required packages.    
```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries. 
```{r}
library(tidyverse)
library(ggplot2)
library(readxl)
library(cowplot)
library(lme4)
library(lmerTest)
library(car)
library(effects)
library(emmeans)
```

# Load data 

```{r}
#read in files
data <- read_csv("data/resazurin/resazurin_data.csv")

data<-data%>%
  pivot_longer(names_to="time", values_to="fluorescence", cols=`0`:`4`)%>%
  mutate(time=as.numeric(time))
```

Read in survival data 
```{r}
surv<-read_excel("data/survival/survival_resazurin.xlsx")

surv<-surv%>%
  pivot_longer(names_to="time", values_to="status", cols=`0`:`24`)%>%
  mutate(time=as.numeric(time))%>%
  rename(date=date.resazurin, sample=oyster)%>%
  mutate(sample=as.character(sample))
```

Combine data
```{r}
data<-left_join(data, surv)
```

Add in final mortality status
```{r}
final<-surv%>%
  filter(time==24)%>%
  rename(final.mortality=status)%>%
  select(!time)

data<-left_join(data, final)
```

We now have resazurin and survival data in the data frame.  

```{r}
data<-data%>%
  mutate(final.mortality = case_when(
    final.mortality == 0 ~ "alive",
    final.mortality == 1 ~ "dead",
    TRUE ~ as.character(final.mortality)  # To handle any other values
  ))
```

# Data preparation 

Plot the data. 

```{r}
data%>%
  ggplot(aes(x=time, y=fluorescence, colour=temperature, group=sample))+
  facet_wrap(~bag*date)+
  geom_point()+
  geom_line()+
  theme_classic()
```

Calculate fluorescence at each time point normalized to the starting value at time 0. 

```{r}
data<-data%>%
  group_by(date, bag, sample)%>%
  arrange(date, bag, sample)%>%
  mutate(fluorescence.norm=fluorescence/first(fluorescence))
```

Plot again 

```{r}
data%>%
  ggplot(aes(x=time, y=fluorescence.norm, colour=temperature, group=sample))+
  facet_wrap(~bag*date)+
  geom_point()+
  geom_line()+
  theme_classic()
```

View blanks

```{r}
data%>%
  filter(type=="blank")%>%
  ggplot(aes(x=time, y=fluorescence.norm, colour=temperature, group=sample))+
  facet_wrap(~bag*date)+
  geom_point()+
  geom_line()+
  theme_classic()
```

Calculate mean change in blank at each time point. 

```{r}
blanks<-data%>%
  filter(type=="blank")%>%
  group_by(date, bag, temperature, time)%>%
  summarise(mean_blank=mean(fluorescence.norm))
```

View summarized blank data. 

```{r}
blanks%>%
  ggplot(aes(x=time, y=mean_blank, colour=temperature))+
  facet_wrap(~bag*date)+
  geom_point()+
  geom_line()+
  theme_classic()
```

Subtract blank values from fluorescence values for oysters. 

```{r}
data<-left_join(data, blanks)

data<-data%>%
  filter(!type=="blank")%>%
  mutate(value=fluorescence.norm-mean_blank)
```

Plot again. 

```{r}
data%>%
  ggplot(aes(x=time, y=fluorescence.norm, colour=temperature, group=sample))+
  facet_wrap(~bag*date*hardening)+
  geom_point()+
  geom_line()+
  theme_classic()
```

Remove unnecessary columns. 

```{r}
data<-data%>%
  select(!type)%>%
  select(!fluorescence.norm)%>%
  select(!mean_blank)%>%
  select(!fluorescence)
```

Plot against final mortality. 

```{r}
data%>%
  filter(time==4)%>%
  filter(temperature=="42C")%>%
  
  ggplot(aes(x=final.mortality, y=value, colour=final.mortality))+
  geom_point()+
  theme_classic()
```

Under temperature stress, there appears to be a positive relationship between fluorescence at 4 hrs and final mortality status.  

# Analysis 

## Analyze treatment effects 

First, log transform the data. 

```{r}
data$log_value<-log10(data$value+1)
```

```{r}
data%>%
  ggplot(aes(x=time, y=log_value, colour=temperature, group=sample))+
  facet_wrap(~bag*date*hardening)+
  geom_point()+
  geom_line()+
  theme_classic()
```

First try a linear mixed effect model and come back to this later for non linear models.  

Will need to come back to random effects here. 
```{r}
model<-lmer(log_value ~ time * temperature * hardening + (1|bag) + (1|date:bag:sample), data=data)

summary(model)
anova(model)
rand(model)

qqPlot(residuals(model))

plot(Effect(c("temperature"), model, transformation=list(link=log)))
plot(Effect(c("hardening"), model, transformation=list(link=log)))
plot(Effect(c("time"), model, transformation=list(link=log)))
plot(Effect(c("time", "temperature"), model, transformation=list(link=log)))
plot(Effect(c("time", "hardening"), model, transformation=list(link=log)))
plot(Effect(c("time", "hardening", "temperature"), model, transformation=list(link=log)))
```

Come back to normality assessment.  

Significant effects of time x temperature x hardening, indicating different slopes in metabolic rate between temperatures modulated by hardening treatment. 

Analyze just with the ending value. 

```{r}
final<-data%>%
  filter(time==4)

model2<-lmer(log_value ~ temperature * hardening + (1|bag:hardening), data=final)

summary(model2)
anova(model2)
rand(model2)

qqPlot(residuals(model2))

plot(Effect(c("hardening", "temperature"), model2, transformation=list(link=log)))
```

Run a model including time for those that lived and died separately 

```{r}
dead<-data%>%
  filter(final.mortality=="dead")

alive<-data%>%
  filter(final.mortality=="alive")
```

Alive
```{r}
model2a<-lmer(log_value ~ time * temperature * hardening + (1|bag:hardening:sample) + (1|bag), data=alive)

summary(model2a)
anova(model2a)
rand(model2a)

qqPlot(residuals(model2a))
```

Plot this. 
```{r}
plot_alive<-alive%>%
  
  ggplot(aes(x=time, y=log_value, color=temperature))+
  facet_grid(~hardening)+
  geom_point(alpha=0.5)+
  geom_smooth(aes(group=temperature))+
  scale_colour_manual(values=c("cyan4", "orange"))+
  theme_classic()+
  xlab("Hour")+
  ggtitle("Oysters that lived");plot_alive
  
ggsave(plot_alive, filename="figures/resazurin/alive_metabolism.png", width=10, height=5)
```

Dead
```{r}
model2b<-lmer(log_value ~ time * temperature * hardening + (1|bag:hardening:sample) + (1|bag), data=dead)

summary(model2b)
anova(model2b)
rand(model2b)

qqPlot(residuals(model2b))

plot(Effect(c("time", "hardening", "temperature"), model2b, transformation=list(link=log)))

emm<-emmeans(model2b, ~hardening|time)
pairs(emm)
```

Plot this. 

```{r}
plot_dead<-dead%>%
  
  ggplot(aes(x=time, y=log_value, color=temperature))+
  facet_grid(~hardening)+
  geom_point(alpha=0.5)+
  geom_smooth(aes(group=temperature))+
  scale_colour_manual(values=c("cyan4", "orange"))+
  theme_classic()+
  xlab("Hour")+
  ggtitle("Oysters that died");plot_dead
  
ggsave(plot_dead, filename="figures/resazurin/dead_metabolism.png", width=10, height=5)
```

## Include mortality in models 

Next, include mortality as a factor. 

```{r}
model3<-lmer(log_value ~ temperature * hardening * final.mortality + (1|bag:hardening), data=final)

summary(model3)
anova(model3)
rand(model3)

qqPlot(residuals(model3))

plot(Effect(c("hardening", "temperature"), model3, transformation=list(link=log)))
plot(Effect(c("final.mortality"), model3, transformation=list(link=log)))
plot(Effect(c("temperature"), model3, transformation=list(link=log)))
```

This shows an interactive effect of hardening and temperature as well as a main effect of mortality and temperature. Rates are lower at 42C, but higher in those that die. Rates are lower in some hardening treatments at high temperature. 

View posthoc. 

```{r}
emm<-emmeans(model3, ~ hardening | temperature | final.mortality)
pairs(emm)

emm<-emmeans(model3, ~ temperature | hardening | final.mortality)
pairs(emm)

emm<-emmeans(model3, ~ final.mortality | hardening | temperature)
pairs(emm)
```

All hardening treatments have lower metabolism at high temperature. The decrease is stronger in immune, fresh water, and fresh water temperature treatments as compared to control and temperature hardening.  

No difference in metabolic rates between temperatures in those that lived in control. There was a difference for all other treatments. 

There was a difference in metabolic rates between temperature treatments for those that died in all treatments EXCEPT temperature hardening. 

## Mortality and rates 

Analyze the relationship between rate and mortality. 

```{r}
model4<-lm(log_value ~ status, data=data)
summary(model4)
```

```{r}
data %>%
  
  ggplot(aes(x=as.factor(status), y=log_value))+
  geom_boxplot()+
  geom_point()+
  theme_classic()
```

Those that died have a higher metabolic rate. 

# Plot data

Look at data to remove outliers if needed. 
```{r}
hist(final$log_value)
```
No obvious outliers.  

Plot final metabolic rates across hardening and temperature treatments for those that lived vs those that died. 

```{r}
final %>%
  
  ggplot(aes(x=hardening, y=log_value, colour=temperature))+
  facet_wrap(~final.mortality)+
  geom_boxplot()+
  geom_point(position=position_jitterdodge(0.1))+
  scale_colour_manual(values=c("cyan4", "orange"))+
  theme_classic()+
  theme(
    axis.text.x=element_text(angle=45, hjust=1, vjust=1)
  )
```

Plot as mean and standard error. 

```{r}
plot1<-final %>%
  group_by(hardening, temperature, final.mortality)%>%
  summarise(mean=mean(log_value), se=sd(log_value)/sqrt(length(log_value)))%>%
  
  ggplot(aes(x=hardening, y=mean, colour=temperature))+
  facet_wrap(~final.mortality)+
  geom_point(position=position_dodge(0.2))+
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se), position=position_dodge(0.2), width=0.4)+
  scale_colour_manual(values=c("cyan4", "orange"))+
  theme_classic()+
  ylim(0,1)+
  ylab("Fluorescence [4 hr]")+
  theme(
    axis.text.x=element_text(angle=45, hjust=1, vjust=1),
    panel.border=element_rect(fill=NA,color="black", size=0.5, 
                                    linetype="solid")
  );plot1

ggsave(plot1, filename="figures/resazurin/resazurin_plot1.png", width=7, height=5)
```

Plot as mean and standard error to focus on mortality differences. 

```{r}
plot2<-final %>%
  group_by(hardening, temperature, final.mortality)%>%
  summarise(mean=mean(log_value), se=sd(log_value)/sqrt(length(log_value)))%>%
  
  ggplot(aes(x=hardening, y=mean, colour=final.mortality))+
  facet_wrap(~temperature)+
  geom_point(position=position_dodge(0.2))+
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se), position=position_dodge(0.2), width=0.4)+
  scale_colour_manual(values=c("gray", "darkred"))+
  theme_classic()+
  ylab("Fluorescence [4 hr]")+
  ylim(0,1)+
  theme(
    axis.text.x=element_text(angle=45, hjust=1, vjust=1),
    panel.border=element_rect(fill=NA,color="black", size=0.5, 
                                    linetype="solid")
  );plot2

ggsave(plot2, filename="figures/resazurin/resazurin_plot2.png", width=7, height=5)
```

```{r}
plot3<-final %>%
  group_by(hardening, temperature, final.mortality)%>%
  summarise(mean=mean(log_value), se=sd(log_value)/sqrt(length(log_value)))%>%
  
  ggplot(aes(x=temperature, y=mean, colour=final.mortality))+
  facet_grid(~hardening)+
  geom_point(position=position_dodge(0.2))+
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se), position=position_dodge(0.2), width=0.4)+
  scale_colour_manual(values=c("gray", "darkred"))+
  theme_classic()+
  geom_line(aes(group=final.mortality), position=position_dodge(0.1))+
  ylab("Fluorescence [4 hr]")+
  ylim(0,1)+
  theme(
    axis.text.x=element_text(angle=45, hjust=1, vjust=1), 
    panel.border=element_rect(fill=NA,color="black", size=0.5, 
                                    linetype="solid")
    );plot3

ggsave(plot3, filename="figures/resazurin/resazurin_plot3.png", width=10, height=4)
```

