---
title: 10K seed resazurin survival analysis
author: "AS Huffmyer"
date: '2024'
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_depth: 6
    toc_float: yes
  pdf_document:
    keep_tex: yes
editor_options: 
  chunk_output_type: console
---


# Set up 

Set up workspace, set options, and load required packages.    
```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries. 
```{r}
library(tidyverse)
library(ggplot2)
library(survival)
library(readxl)
library(ggsurvfit)
library(gtsummary)
library(cardx)
library(cowplot)
```

# Read data 

Read in data. 
```{r}
data<-read_excel("data/survival/survival_resazurin.xlsx")
```

Turn into long format. 
```{r}
data<-data%>%
  pivot_longer(names_to="time", values_to="status", cols=`0`:`24`)%>%
  mutate(time=as.numeric(time))
```

Create data frames with control and high temperature subsets. 
```{r}
control<-data%>%filter(temperature=="18C")
stress<-data%>%filter(temperature=="42C")
```


# Generate Kaplan Meier survival curves 

## Control temperature 18C

```{r}
s1 <- survfit(Surv(time, status) ~ hardening, data = control)
str(s1)
```

Plot the survival function
```{r}
survfit2(Surv(time, status) ~ hardening+temperature, data = control) %>% 
  ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability"
  )
```

Estimate survival at 24 hours. 
```{r}
summary(survfit(Surv(time, status) ~ hardening, data = control), times = 24)
```

Estimated control survival is ~65% in control temperature. 

Use a log rank model to determine statistical differences in curves. 
```{r}
survdiff(Surv(time, status) ~ hardening, data = control)

# Call:
# survdiff(formula = Surv(time, status) ~ hardening, data = control)
# 
#                                     N Observed Expected (O-E)^2/E (O-E)^2/V
# hardening=control                 360       18       18    0.0000     0.000
# hardening=fresh-water             240       12       12    0.0000     0.000
# hardening=fresh-water-temperature 120        6        6    0.0000     0.000
# hardening=immune                  240       13       12    0.0833     0.133
# hardening=temperature             240       11       12    0.0833     0.133
# 
#  Chisq= 0.2  on 4 degrees of freedom, p= 1 
```
There is no difference in survival at 18C at p=1. 

Analyze again with a Cox proportional hazards model. 
```{r}
coxph(Surv(time, status) ~ hardening, data = control)

# Call:
# coxph(formula = Surv(time, status) ~ hardening, data = control)
# 
#                                      coef exp(coef) se(coef)      z     p
# hardeningfresh-water              0.01956   1.01975  0.37269  0.052 0.958
# hardeningfresh-water-temperature  0.03148   1.03198  0.47142  0.067 0.947
# hardeningimmune                   0.07701   1.08005  0.36398  0.212 0.832
# hardeningtemperature             -0.06831   0.93397  0.38272 -0.178 0.858
# 
# Likelihood ratio test=0.13  on 4 df, p=0.9979
# n= 1200, number of events= 60  

coxph(Surv(time, status) ~ hardening, data = control) %>% 
  tbl_regression(exp = TRUE) 
```
Control is the reference level. No differences in survival at 18C. 

# Generate Kaplan Meier survival curves 

## Stress temperature 42C

```{r}
s2 <- survfit(Surv(time, status) ~ hardening, data = stress)
str(s2)
```

Plot the survival function
```{r}
survfit2(Surv(time, status) ~ hardening+temperature, data = stress) %>% 
  ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability"
  )
```

Estimate survival at 24 hours. 
```{r}
summary(survfit(Surv(time, status) ~ hardening, data = stress), times = 24)
```

Estimated stress survival is 2-8%. Temperature hardened is the lowest.    

Use a log rank model to determine statistical differences in curves. 
```{r}
survdiff(Surv(time, status) ~ hardening, data = stress)

# Call:
# survdiff(formula = Surv(time, status) ~ hardening, data = stress)
# 
#                                     N Observed Expected (O-E)^2/E (O-E)^2/V
# hardening=control                 360      116    133.5     2.294     5.558
# hardening=fresh-water             240       83     89.0     0.404     0.858
# hardening=fresh-water-temperature 120       40     44.5     0.455     0.858
# hardening=immune                  240       97     89.0     0.719     1.525
# hardening=temperature             240      109     89.0     4.494     9.529
# 
#  Chisq= 14.2  on 4 degrees of freedom, p= 0.007
```
There is a significant difference in survival at p=0.007. 

Analyze again with a Cox proportional hazards model. 
```{r}
coxph(Surv(time, status) ~ hardening, data = stress)

# Call:
# coxph(formula = Surv(time, status) ~ hardening, data = stress)
# 
#                                     coef exp(coef) se(coef)     z      p
# hardeningfresh-water             0.08467   1.08835  0.14377 0.589 0.5559
# hardeningfresh-water-temperature 0.03658   1.03725  0.18336 0.199 0.8419
# hardeningimmune                  0.20795   1.23115  0.13764 1.511 0.1308
# hardeningtemperature             0.41045   1.50749  0.13344 3.076 0.0021
# 
# Likelihood ratio test=10.75  on 4 df, p=0.02958
# n= 1200, number of events= 445 

coxph(Surv(time, status) ~ hardening, data = stress) %>% 
  tbl_regression(exp = TRUE) 
```
Control is the reference level. There is a significant difference between control and temperature hardening survival. Marginal difference between control and fresh water.   

# Generate plots 
 
Set theme. 
```{r}
my_theme<-theme_classic()
```

Control
```{r}
plot1<-survfit2(Surv(time, status) ~ hardening, data = control) %>% 
  ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title="18°C"
  )+
  ylim(0,1)+
  my_theme+
  geom_text(x=10, y=0.2, label="Cox PH p=0.997")+
  theme(legend.position="none")
```

Stress
```{r}
plot2<-survfit2(Surv(time, status) ~ hardening, data = stress) %>% 
  ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title = "42°C"
  )+
  ylim(0,1)+
  my_theme+
  geom_text(x=10, y=0.2, label="Cox PH p=0.030")+
  geom_text(x=10, y=0.15, label="Temp vs control p=0.002")+
  theme(legend.position="right")
```

Assemble plot
```{r}
plots<-plot_grid(plot1, plot2, rel_widths=c(0.65,1), ncol=2)

ggsave(plots, filename="figures/survival/KMcurves.png", width=10, height=4)
```


