---
title: 10K seed resazurin survival analysis
author: "AS Huffmyer"
date: '2024'
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_depth: 6
    toc_float: yes
  pdf_document:
    keep_tex: yes
editor_options: 
  chunk_output_type: console
---


# Set up 

Set up workspace, set options, and load required packages.    
```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries. 
```{r}
library(tidyverse)
library(ggplot2)
library(survival)
library(readxl)
library(ggsurvfit)
library(gtsummary)
library(cardx)
library(cowplot)
```

# Read data 

Read in data. 
```{r}
data<-read_excel("data/survival/survival_resazurin.xlsx")
```

Turn into long format. 
```{r}
data<-data%>%
  pivot_longer(names_to="time", values_to="status", cols=`0`:`24`)%>%
  mutate(time=as.numeric(time))
```

Create data frames with control and high temperature subsets. 
```{r}
control<-data%>%filter(temperature=="18C")
stress<-data%>%filter(temperature=="42C")
```


# Generate Kaplan Meier survival curves 

## Control temperature 18C

```{r}
s1 <- survfit(Surv(time, status) ~ hardening, data = control)
str(s1)
```

Plot the survival function
```{r}
survfit2(Surv(time, status) ~ hardening+temperature, data = control) %>% 
  ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability"
  )
```

Estimate survival at 24 hours. 
```{r}
summary(survfit(Surv(time, status) ~ hardening, data = control), times = 24)
```

Estimated control survival is ~75% in control temperature, with fresh water as the lowest mean.    

Use a log rank model to determine statistical differences in curves. 
```{r}
survdiff(Surv(time, status) ~ hardening, data = control)

# Call:
# survdiff(formula = Surv(time, status) ~ hardening, data = control)
# 
#                                     N Observed Expected (O-E)^2/E (O-E)^2/V
# hardening=control                 240       10    11.50    0.1957    0.3369
# hardening=fresh-water             240       12    11.50    0.0217    0.0374
# hardening=fresh-water-temperature 120        6     5.75    0.0109    0.0160
# hardening=immune                  120        7     5.75    0.2717    0.4010
# hardening=temperature             240       11    11.50    0.0217    0.0374
# 
#  Chisq= 0.7  on 4 degrees of freedom, p= 1
```
There is no difference in survival at 18C at p=1. 

Analyze again with a Cox proportional hazards model. 
```{r}
coxph(Surv(time, status) ~ hardening, data = control)

# Call:
# coxph(formula = Surv(time, status) ~ hardening, data = control)
# 
#                                    coef exp(coef) se(coef)     z     p
# hardeningfresh-water             0.2085    1.2318   0.4282 0.487 0.626
# hardeningfresh-water-temperature 0.2213    1.2477   0.5164 0.429 0.668
# hardeningimmune                  0.3263    1.3858   0.4928 0.662 0.508
# hardeningtemperature             0.1210    1.1286   0.4369 0.277 0.782
# 
# Likelihood ratio test=0.52  on 4 df, p=0.9716
# n= 960, number of events= 46  

coxph(Surv(time, status) ~ hardening, data = control) %>% 
  tbl_regression(exp = TRUE) 
```
Control is the reference level. No differences in survival at 18C. 

# Generate Kaplan Meier survival curves 

## Stress temperature 42C

```{r}
s2 <- survfit(Surv(time, status) ~ hardening, data = stress)
str(s2)
```

Plot the survival function
```{r}
survfit2(Surv(time, status) ~ hardening+temperature, data = stress) %>% 
  ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability"
  )
```

Estimate survival at 24 hours. 
```{r}
summary(survfit(Surv(time, status) ~ hardening, data = stress), times = 24)
```

Estimated stress survival is 5-20%. Temperature hardened is the lowest.    

Use a log rank model to determine statistical differences in curves. 
```{r}
survdiff(Surv(time, status) ~ hardening, data = stress)

# Call:
# survdiff(formula = Surv(time, status) ~ hardening, data = stress)
# 
#                                     N Observed Expected (O-E)^2/E (O-E)^2/V
# hardening=control                 240       63       84    5.2500   12.0917
# hardening=fresh-water             240       83       84    0.0119    0.0274
# hardening=fresh-water-temperature 120       40       42    0.0952    0.1880
# hardening=immune                  120       41       42    0.0238    0.0470
# hardening=temperature             240      109       84    7.4405   17.1367
# 
#  Chisq= 22.1  on 4 degrees of freedom, p= 2e-04 
```
There is a significant difference in survival at p<0.001. 

Analyze again with a Cox proportional hazards model. 
```{r}
coxph(Surv(time, status) ~ hardening, data = stress)

# Call:
# coxph(formula = Surv(time, status) ~ hardening, data = stress)
# 
#                                    coef exp(coef) se(coef)     z        p
# hardeningfresh-water             0.3252    1.3843   0.1671 1.946   0.0517
# hardeningfresh-water-temperature 0.2764    1.3184   0.2022 1.367   0.1716
# hardeningimmune                  0.2754    1.3170   0.2007 1.372   0.1700
# hardeningtemperature             0.6526    1.9205   0.1584 4.120 3.79e-05
# 
# Likelihood ratio test=18.11  on 4 df, p=0.001177
# n= 960, number of events= 336 

coxph(Surv(time, status) ~ hardening, data = stress) %>% 
  tbl_regression(exp = TRUE) 
```
Control is the reference level. There is a significant difference between control and temperature hardening survival. Marginal difference between control and fresh water.   

# Generate plots 
 
Set theme. 
```{r}
my_theme<-theme_classic()
```

Control
```{r}
plot1<-survfit2(Surv(time, status) ~ hardening, data = control) %>% 
  ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title="18°C"
  )+
  ylim(0,1)+
  my_theme+
  geom_text(x=10, y=0.2, label="Cox PH p=0.70")+
  theme(legend.position="none")
```

Stress
```{r}
plot2<-survfit2(Surv(time, status) ~ hardening, data = stress) %>% 
  ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title = "42°C"
  )+
  ylim(0,1)+
  my_theme+
  geom_text(x=10, y=0.2, label="Cox PH p=0.001")+
  geom_text(x=10, y=0.15, label="Temp vs control p<0.001")+
  theme(legend.position="right")
```

Assemble plot
```{r}
plots<-plot_grid(plot1, plot2, rel_widths=c(0.65,1), ncol=2)

ggsave(plots, filename="figures/survival/KMcurves.png", width=10, height=4)
```


